Vectorizing Python Constructs Using Pythran and Boost SIMD

### 2013-11-27 doc/papers/wpmvp14/paper.tex
Piece of Boost.SIMD

@@

\authorinfo{Serge Guelton}
           {QuarksLab, T{\'e}l{\'e}com Bretagne}
           {sguelton@quarkslab.com}
\authorinfo{Jo{\"e}l Falcou}
           {MetaScale}
           {joel.falcou@metascale.org}
\authorinfo{Pierrick Brunet}
           {INRIA/MOAIS}
           {pierrick.brunet@inria.fr}

@@

%%
%%
\section{Vectorization of a Meta-Program}
\label{sec:meta-vectorization}

\subsection{Boost SIMD}

>>>>

\providecommand{\boostsimd}{\textsc{Boost.SIMD}}
\providecommand{\cpp}[1][~]{\textsc{C++}#1}
\providecommand{\ie}[1][~]{\textit{i.e.}#1}
\providecommand{\eg}[1][~]{\textit{e.g.#1}}

@@

\authorinfo{Serge Guelton}
           {QuarksLab, T{\'e}l{\'e}com Bretagne}
           {sguelton@quarkslab.com}
\authorinfo{Jo{\"e}l Falcou}
           {LRI, Universit\'e Paris-Sud}
           {joel.falcou@lri.fr}
\authorinfo{Pierrick Brunet}
           {INRIA/MOAIS}
           {pierrick.brunet@inria.fr}

@@

%%
%%
\section{Vectorization of a Meta-Program}
\label{sec:meta-vectorization}
SIMD extensions have been a feature of choice for processor manufacturers for a
couple of decades. However, programming applications that take advantage of the 
SIMD extension available on the current target remains a complex task. Programmers 
that use low-level intrinsics have to deal with a verbose programming style due 
to the fact that SIMD instructions sets cover a few common functionnalities, requiring
to bury the initial algorithms in architecture specific implementation details.
Furthermore, these efforts have to be repeated for every different extension
that one may want to support, making design and maintenance of such applications 
very time consuming.

\boostsimd{} is a high-level C++ library to program SIMD architectures. 
Designed as an \textit{Embedded Domain Specific Language}, \boostsimd{} 
provides both expresiveness and performance by using generic programming 
to handle vectorization. \boostsimd{} does not only provide a portable 
way but also enable the use of common programming idioms when designing 
SIMD-aware algorithms. Two components are provided:

\begin{itemize}
\item \textbf{An abstraction of SIMD registers} for portable algorithm design,
coupled with a large set of functions covering the classical set of operators
along with a sensible amount (200+) of mathematical functions and utility functions,
\item \textbf{A support for \cpp standard components} like \texttt{Iterator}
over SIMD \texttt{Range}, SIMD-aware allocators and SIMD-aware STL algorithms.
\end{itemize}

\subsection{SIMD register abstraction}
The first level of abstraction introduced by \textsc{Boost.} \textsc{SIMD} is
the \texttt{pack} class. For a given type \texttt{T} and a given static integral 
value \texttt{N} (\texttt{N} being a power of 2), a \texttt{pack} encapsulates
the best type able to store a sequence of \texttt{N} elements of type \texttt{T}.
For arbitrary \texttt{T} and \texttt{N}, this type is simply \texttt{std::array<T,N>}
but when \texttt{T} and \texttt{N} matches the type and width of a SIMD register,
the architecture-specific type used to represent this register is used instead.
This semantic provides a way to use arbitrarily large SIMD registers on any system
and let the library select the best vectorizable type to handle them. By default,
if \texttt{N} is not provided, \texttt{pack} will automatically select a value
that will trigger the selection of the native SIMD register type. Moreover, by
carrying informations about its underlying scalar type, \texttt{pack} enables
proper instruction selection even when used on extensions (like SSE2 and above)
that map all integral type to a single SIMD type (\texttt{\_\_m128i} for SSE2).\\

\texttt{pack} handles these low-level SIMD register type as regular objects with
value semantics, which includes the ability to be constructed or copied from a
single scalar value, list of scalar values, iterator or range. In each case, the
proper register loading strategy (splat, set, load or gather) will be issued.
\texttt{pack} also takes care of issues like boolean predicates support and provides
both a range and a tuple-like interface.

\subsection{Interaction with Standard Algorithm}
Modern \cpp programming style based on Generic Programming usually leads to an
intensive use of various STL components like Iterators. \boostsimd{} provides
iterator adaptors that turn regular random access iterators into iterators
suitable for SIMD processing. These adaptors act as free functions taking
regular iterators as parameters and return iterators that output \texttt{pack}
whenever dereferenced. These iterators are then usable directly in usual STL
algorithms such as \texttt{transform} or \texttt{fold} (Listing \ref{stl_example}).

\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={SIMD Iterator with STL algorithm},label=stl_example]
std::vector<int,allocator<int> > v(N),r(N);

std::transform ( simd::begin(v.begin())
               , simd::end(v.end())
               , simd::begin(r.begin())
               , [](pack<int> const& p)
                 { 
                   return p*p; 
                 } 
              );
\end{lstlisting}

Previous examples of integration within standard algorithm are still limited.
Applying \texttt{transform} or \texttt{fold} algorithm on SIMD aware data
requires the size of the data to be an exact multiple of the SIMD register width
and, in the case of fold, to potentially perform additional operations at the
end of its call. To alleviate this limitation, \boostsimd{}
provides its own overload for both \texttt{transform} and \texttt{fold} that take
care of potential trailing data and performs proper completion of \texttt{fold}.
Moreover, as \boostsimd{} provides default implementation of all its function for scalar values,
we can rely on polymorphic callable object to write a mixed scalar/SIMD call to algortihm
like \texttt{transform}. By using the \texttt{boost::simd::transform} algorithm, code
handlign alignment, scalar prologue and epilogue is generated.

\begin{lstlisting}[language=c++, label={lst:square-cxx}, caption={Polymorphic SIMD algorithm call},label=stl_poly]
std::vector<int,allocator<int> > v(128), r(128);

struct f
{
  template<class T>
  auto operator()(T const& p)
  -> decltype(p * p)
  {
    return p * p;
  }
};

simd::transform( v.begin(), v.end()
               , r.begin()
               , f() 
               );
\end{lstlisting}

Code written this way keeps a conventional structure and facilitate an usage of template
functors for both scalar and SIMD cases also helps maximizing code reuse.
